---
title: "Tidy Data Extraction from Seurat Objects"
description: >
  Extract cell-level metadata, embeddings, and feature expression data as tidy
  tibbles — replacing Seurat's FetchData() with explicit, composable functions.
---

```{r}
#| label: setup
#| include: false
library(BadranSeq)
library(Seurat)
library(dplyr)
library(ggplot2)
data(pbmc3k, package = "BadranSeq")
```

## Motivation

Seurat's `FetchData()` is the standard way to pull data out of a Seurat object
for custom plotting or downstream analysis. It works, but has a few rough edges:

- Returns a `data.frame` with cell barcodes as **rownames**, not a column
- Mixes expression data, embeddings, and metadata into one flat call
- No control over which layer (counts vs normalised) to extract
- Returns wide format for features, which requires manual pivoting for tidy
  workflows

BadranSeq provides two focused alternatives:

| Function | Returns | Use case |
|----------|---------|----------|
| `fetch_cell_data()` | One row per cell | Metadata, embeddings — no expression |
| `fetch_feature_data()` | One row per cell × feature | Expression data + optional metadata/embeddings |

Both return tibbles with `cell_id` as an explicit column.

## The FetchData problem

Here is what `Seurat::FetchData()` returns for a simple request — UMAP
coordinates plus a gene:

```{r}
#| label: fetchdata-seurat
head(Seurat::FetchData(pbmc3k, vars = c("umap_1", "umap_2", "CD3D")))
```

Cell barcodes are trapped in rownames. Expression and embeddings are mingled.
There is no way to request a specific layer.

Compare with BadranSeq's approach — embeddings and expression are requested
through separate, explicit arguments:

```{r}
#| label: fetchdata-badranseq
fetch_feature_data(
  pbmc3k,
  features = "CD3D",
  layer = "data",
  reductions = "umap",
  metadata = FALSE
)
```

`cell_id` is a column. The feature name is explicit. The layer is explicit. The
output is a tibble ready for `ggplot2` or `dplyr` pipelines.

## Cell-level data

`fetch_cell_data()` extracts metadata and embeddings — no expression data. This
is the right tool when you need cell annotations for plotting or filtering.

### All metadata (default)

```{r}
#| label: cell-all-meta
fetch_cell_data(pbmc3k)
```

### Add UMAP coordinates

```{r}
#| label: cell-umap
fetch_cell_data(pbmc3k, reductions = "umap")
```

### Select specific metadata columns

```{r}
#| label: cell-specific-meta
fetch_cell_data(
  pbmc3k,
  reductions = "umap",
  metadata = c("seurat_clusters", "nFeature_RNA")
)
```

### Embeddings only (no metadata)

```{r}
#| label: cell-embed-only
fetch_cell_data(pbmc3k, reductions = c("pca", "umap"), metadata = FALSE)
```

## Feature expression data

`fetch_feature_data()` returns one row per cell per feature in long format.
Expression values are columns named after the requested layers.

### Single gene, normalised expression

```{r}
#| label: feat-single
fetch_feature_data(pbmc3k, features = "CD3D")
```

### Multiple genes

The `feature` column is a factor preserving your input order — useful for
controlling facet ordering in plots:

```{r}
#| label: feat-multi
fetch_feature_data(
  pbmc3k,
  features = c("CD3D", "CD8A", "CD14", "MS4A1"),
  metadata = "seurat_clusters"
)
```

### Multiple layers side-by-side

Request both raw counts and normalised data. Each layer becomes its own column:

```{r}
#| label: feat-layers
fetch_feature_data(
  pbmc3k,
  features = "CD3D",
  layer = c("counts", "data"),
  metadata = FALSE
)
```

### With embeddings for custom plots

Combine expression with UMAP coordinates for a fully self-contained plotting
tibble:

```{r}
#| label: feat-comparison
cd3d <- fetch_feature_data(
  pbmc3k,
  features = "CD3D",
  reductions = "umap",
  metadata = FALSE
)

ggplot(cd3d, aes(x = UMAP1, y = UMAP2, colour = data)) +
  geom_point(size = 0.5) +
  scale_colour_viridis_c(option = "B", direction = -1) +
  theme_minimal() +
  labs(colour = "CD3D")
```

Compare this with the equivalent using `FetchData()`:

```{r}
#| label: feat-comparison-seurat
fd <- Seurat::FetchData(pbmc3k, vars = c("umap_1", "umap_2", "CD3D"))

ggplot(fd, aes(x = umap_1, y = umap_2, colour = CD3D)) +
  geom_point(size = 0.5) +
  scale_colour_viridis_c(option = "B", direction = -1) +
  theme_minimal()
```

Both produce the same plot, but `fetch_feature_data()` gives you a tidy tibble
where the feature name is data, not a column name — making it straightforward to
facet across multiple genes without reshaping.

## PCA diagnostics

BadranSeq includes two functions for understanding PCA results:
`get_pca_variance()` for the raw numbers and `EnhancedElbowPlot()` for the
visual.

### Variance explained table

```{r}
#| label: pca-variance
get_pca_variance(pbmc3k, n_pcs = 10)
```

The cumulative variance column helps decide how many PCs to retain — a common
question when choosing dimensions for clustering and UMAP.

### Enhanced elbow plot

`EnhancedElbowPlot()` improves on Seurat's `ElbowPlot()` by optionally showing
a variance percentage axis and a visual cutoff line:

```{r}
#| label: elbow-plot
#| fig.width: 7
#| fig.height: 5
EnhancedElbowPlot(pbmc3k, ndims = 30, cutoff_pc = 10)
```

The red dashed line at PC 10 marks the analyst's chosen cutoff — a quick
visual reference when deciding how many PCs to retain for downstream analysis.

## Putting it together

A practical example: extract expression for T cell markers with cluster
annotations, then produce a grouped violin plot entirely from the tidy tibble.

```{r}
#| label: workflow
#| fig.width: 8
#| fig.height: 5
markers <- fetch_feature_data(
  pbmc3k,
  features = c("CD3D", "CD8A", "CD4", "IL7R"),
  metadata = "seurat_clusters"
)

ggplot(markers, aes(x = seurat_clusters, y = data, fill = seurat_clusters)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~feature, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Cluster", y = "Normalised expression")
```

Because `feature` is a factor in input order, the facet panels appear in the
biologically meaningful order you specified — not alphabetically.
