---
title: "Publication-Ready Single-Cell Visualisation"
description: >
  Generate publication-quality dimensionality reduction plots with sensible
  defaults — cell borders, cluster labels, variance annotations, and
  split-panel silhouettes out of the box.
---

```{r}
#| label: setup
#| include: false
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)
data(pbmc3k, package = "BadranSeq")
```

## Motivation

Seurat's plotting functions are designed for exploration. Getting them to
publication quality typically requires layering `ggplot2` theme adjustments,
adding cell borders, repositioning legends, and reformatting axis labels — the
same boilerplate repeated across every figure in a manuscript.

BadranSeq wraps this into a small set of functions with opinionated defaults:
cell borders on, cluster labels on, clean white theme, legend at bottom. The
goal is a plot that looks good in a paper without additional customisation.

## From Seurat defaults to publication-ready

The quickest way to see the difference — the same data plotted with
`Seurat::DimPlot()` and `BadranSeq::do_UmapPlot()`:

```{r}
#| label: comparison
#| fig.width: 14
#| fig.height: 5
p1 <- Seurat::DimPlot(pbmc3k, reduction = "umap") +
  ggtitle("Seurat::DimPlot()")

p2 <- do_UmapPlot(pbmc3k) +
  ggtitle("BadranSeq::do_UmapPlot()")

p1 + p2
```

BadranSeq adds cell borders, boxed cluster labels, and a clean theme
automatically. No `theme()` calls needed.

## UMAP plots

`do_UmapPlot()` is the workhorse for UMAP visualisation.

### Basic usage

```{r}
#| label: umap-basic
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k)
```

### Group by a metadata column

```{r}
#| label: umap-groupby
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, group.by = "seurat_annotations")
```

### Group by multiple metadata columns

Pass a vector of column names to create side-by-side panels:

```{r}
#| label: umap-multi-groupby
#| fig.width: 14
#| fig.height: 5
do_UmapPlot(pbmc3k, group.by = c("seurat_clusters", "seurat_annotations"))
```

### Control panel layout

When combining multiple panels, use `ncol` and `nrow` to control the
patchwork layout. For example, stack the panels vertically with `ncol = 1`:

```{r}
#| label: umap-layout-control
#| fig.width: 7
#| fig.height: 10
do_UmapPlot(pbmc3k,
            group.by = c("seurat_clusters", "seurat_annotations"),
            ncol = 1)
```

### Hide the legend

```{r}
#| label: umap-no-legend
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, show.legend = FALSE)
```

### Adjust point size and remove labels

```{r}
#| label: umap-customise
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, pt.size = 0.3, label = FALSE)
```

### Remove cell borders

```{r}
#| label: umap-no-borders
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, plot_cell_borders = FALSE)
```

## PCA with automatic variance labels

`do_PcaPlot()` does what Seurat's PCA plot does not — it calculates variance
explained per PC and formats the axis labels as "PC1 (X.X%)".

```{r}
#| label: pca-comparison
#| fig.width: 14
#| fig.height: 5
p1 <- Seurat::DimPlot(pbmc3k, reduction = "pca") +
  ggtitle("Seurat::DimPlot(reduction = 'pca')")

p2 <- do_PcaPlot(pbmc3k) +
  ggtitle("BadranSeq::do_PcaPlot()")

p1 + p2
```

The variance annotation is automatic — no manual calculation or label formatting
required.

### Different dimension pairs

Explore PC2 vs PC3:

```{r}
#| label: pca-dims
#| fig.width: 7
#| fig.height: 5
do_PcaPlot(pbmc3k, dims = c(2, 3))
```

## The unified entry point

`do_DimPlot()` routes to the appropriate handler based on the `reduction`
argument:

| `reduction` | Routes to | Axis labels |
|-------------|-----------|-------------|
| `"umap"` | `do_UmapPlot()` | UMAP1, UMAP2 |
| `"pca"` | `do_PcaPlot()` | PC1 (X.X%), PC2 (X.X%) |
| other | `.do_DimPlot_internal()` | REDUCTION1, REDUCTION2 |

```{r}
#| label: dimplot-routing
#| fig.width: 14
#| fig.height: 5
p1 <- do_DimPlot(pbmc3k, reduction = "umap") + ggtitle("reduction = 'umap'")
p2 <- do_DimPlot(pbmc3k, reduction = "pca") + ggtitle("reduction = 'pca'")

p1 + p2
```

Use `do_DimPlot()` when writing general-purpose code that should work with any
reduction. Use the specific functions (`do_UmapPlot()`, `do_PcaPlot()`) when
you know the reduction type.

## Feature expression plots

`do_FeaturePlot()` overlays gene expression on the embedding using a viridis
colour scale.

```{r}
#| label: featureplot
#| fig.width: 14
#| fig.height: 5
p1 <- do_FeaturePlot(pbmc3k, features = "CD3D")
p2 <- do_FeaturePlot(pbmc3k, features = "CD14")
p3 <- do_FeaturePlot(pbmc3k, features = "MS4A1")

p1 + p2 + p3
```

Cells are ordered by expression by default (`order = TRUE`), so high-expressing
cells are plotted on top — avoiding the common problem of rare populations being
hidden beneath low-expressing cells.

## Split-panel silhouettes

This is BadranSeq's signature visualisation feature. When you pass `split.by`,
each panel shows:

1. A **grey silhouette** of all cells (spatial context)
2. **Black borders** on all cells
3. **Coloured cells** only for the current split category

This preserves the spatial layout across panels, making it easy to see where a
subpopulation sits relative to the full dataset.

First, add a synthetic condition to demonstrate:

```{r}
#| label: split-setup
set.seed(42)
pbmc3k$condition <- sample(
  c("Control", "Treatment"),
  ncol(pbmc3k),
  replace = TRUE
)
```

Now compare Seurat's default split with BadranSeq's silhouette approach:

```{r}
#| label: split-comparison
#| fig.width: 14
#| fig.height: 5
p1 <- Seurat::DimPlot(pbmc3k, reduction = "umap", split.by = "condition") +
  ggtitle("Seurat::DimPlot(split.by)")

p1
```

```{r}
#| label: split-silhouette
#| fig.width: 14
#| fig.height: 7
do_UmapPlot(pbmc3k, split.by = "condition")
```

In the Seurat version, each panel only shows its own cells — you lose the sense
of where those cells sit in the overall structure. In BadranSeq's version, the
grey background preserves this context: you can immediately see which clusters
are enriched in each condition.

The silhouette approach works with any grouping:

```{r}
#| label: split-annotations
#| fig.width: 14
#| fig.height: 7
do_UmapPlot(pbmc3k, group.by = "seurat_annotations", split.by = "condition")
```

## Violin plots

BadranSeq provides two violin plot functions built on
`fetch_feature_data()`:

- **`do_StatsViolinPlot()`** — wraps `ggstatsplot::ggbetweenstats()` for
  between-group comparisons with built-in statistical testing.
- **`do_ViolinPlot()`** — a pure ggplot2 violin with boxplot overlay,
  jittered points, and centrality markers (no statistical testing).

The primary function is `do_StatsViolinPlot()`. If you want the same
aesthetic without statistics, use `do_ViolinPlot()` or set
`pairwise.display = "none"` in `do_StatsViolinPlot()`.

### Statistical violin plots

`do_StatsViolinPlot()` runs a nonparametric omnibus test
(Kruskal–Wallis) and pairwise comparisons (Dunn's test) by default,
annotating the plot with p-values:

```{r}
#| label: stats-violin-basic
#| fig.width: 7
#| fig.height: 6
do_StatsViolinPlot(pbmc3k, features = "CD3D",
                   group.by = "seurat_annotations",
                   pairwise.display = "none")
```

With nine annotation groups, pairwise brackets would overwhelm the
figure. The `pairwise.display` argument controls this:

### Controlling pairwise comparisons

| Value              | Behaviour                          |
|--------------------|------------------------------------|
| `"significant"`    | Only significant pairs (default)   |
| `"non-significant"`| Only non-significant pairs         |
| `"all"`            | Every pairwise comparison          |
| `"none"`           | Omnibus test only, no brackets     |

Setting `pairwise.display = "none"` keeps the subtitle showing the
overall test statistic while removing all bracket annotations — useful
for publication figures where the brackets would overlap.

### Comparing specific groups with `group.levels`

Often you do not want to compare all cell types — perhaps only a few are
biologically relevant. The `group.levels` argument filters the grouping
variable to a specified subset *before* the statistical test is run, so
the test is performed only on the groups you care about.

With fewer groups, pairwise brackets become readable again:

```{r}
#| label: stats-violin-group-levels
#| fig.width: 7
#| fig.height: 6
do_StatsViolinPlot(pbmc3k, features = "CD3D",
                   group.by = "seurat_annotations",
                   group.levels = c("Naive CD4 T", "Memory CD4 T", "CD8 T"))
```

This is cleaner than subsetting the Seurat object manually and avoids
the need to drop unused factor levels.

### Parametric and Bayesian alternatives

The default nonparametric test is appropriate for most scRNA-seq data,
but you can switch to parametric (ANOVA / *t*-test), robust, or Bayesian
tests via the `type` argument:

```{r}
#| label: stats-violin-parametric
#| fig.width: 7
#| fig.height: 6
do_StatsViolinPlot(pbmc3k, features = "CD3D",
                   group.by = "seurat_annotations",
                   group.levels = c("Naive CD4 T", "Memory CD4 T", "CD8 T"),
                   type = "parametric")
```

### Multiple genes

Pass a vector of gene names to get a combined patchwork layout:

```{r}
#| label: stats-violin-multi
#| fig.width: 14
#| fig.height: 6
do_StatsViolinPlot(pbmc3k,
                   features = c("CD3D", "CD8A"),
                   group.by = "seurat_annotations",
                   pairwise.display = "none",
                   ncol = 2)
```

### Plain violin plots

`do_ViolinPlot()` produces a violin + boxplot + jitter + centrality
marker (median diamond) without any statistical annotation. Use this
when you want a clean descriptive plot:

```{r}
#| label: plain-violin
#| fig.width: 7
#| fig.height: 5
do_ViolinPlot(pbmc3k, features = "CD3D",
              group.by = "seurat_annotations")
```

## Sankey / alluvial diagrams

`do_SankeyPlot()` visualises relationships between categorical metadata
variables using an alluvial (Sankey) diagram. Each stratum is labelled
and coloured; flows connect cells across variables.

### Clusters to cell types

A common use case — mapping cluster IDs to biological annotations:

```{r}
#| label: sankey-basic
#| fig.width: 8
#| fig.height: 6
do_SankeyPlot(pbmc3k,
              columns = c("seurat_clusters", "seurat_annotations"))
```

This immediately shows which clusters map cleanly to a single cell type
and which are split across multiple annotations.
