---
title: "Publication-Ready Single-Cell Visualisation"
description: >
  Generate publication-quality dimensionality reduction plots with sensible
  defaults — cell borders, cluster labels, variance annotations, and
  split-panel silhouettes out of the box.
---

```{r}
#| label: setup
#| include: false
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)
data(pbmc3k, package = "BadranSeq")
```

## Motivation

Seurat's plotting functions are designed for exploration. Getting them to
publication quality typically requires layering `ggplot2` theme adjustments,
adding cell borders, repositioning legends, and reformatting axis labels — the
same boilerplate repeated across every figure in a manuscript.

BadranSeq wraps this into a small set of functions with opinionated defaults:
cell borders on, cluster labels on, clean white theme, legend at bottom. The
goal is a plot that looks good in a paper without additional customisation.

## From Seurat defaults to publication-ready

The quickest way to see the difference — the same data plotted with
`Seurat::DimPlot()` and `BadranSeq::do_UmapPlot()`:

```{r}
#| label: comparison
#| fig.width: 14
#| fig.height: 5
p1 <- Seurat::DimPlot(pbmc3k, reduction = "umap") +
  ggtitle("Seurat::DimPlot()")

p2 <- do_UmapPlot(pbmc3k) +
  ggtitle("BadranSeq::do_UmapPlot()")

p1 + p2
```

BadranSeq adds cell borders, boxed cluster labels, and a clean theme
automatically. No `theme()` calls needed.

## UMAP plots

`do_UmapPlot()` is the workhorse for UMAP visualisation.

### Basic usage

```{r}
#| label: umap-basic
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k)
```

### Group by a metadata column

```{r}
#| label: umap-groupby
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, group.by = "seurat_annotations")
```

### Adjust point size and remove labels

```{r}
#| label: umap-customise
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, pt.size = 0.3, label = FALSE)
```

### Remove cell borders

```{r}
#| label: umap-no-borders
#| fig.width: 7
#| fig.height: 5
do_UmapPlot(pbmc3k, plot_cell_borders = FALSE)
```

## PCA with automatic variance labels

`do_PcaPlot()` does what Seurat's PCA plot does not — it calculates variance
explained per PC and formats the axis labels as "PC1 (X.X%)".

```{r}
#| label: pca-comparison
#| fig.width: 14
#| fig.height: 5
p1 <- Seurat::DimPlot(pbmc3k, reduction = "pca") +
  ggtitle("Seurat::DimPlot(reduction = 'pca')")

p2 <- do_PcaPlot(pbmc3k) +
  ggtitle("BadranSeq::do_PcaPlot()")

p1 + p2
```

The variance annotation is automatic — no manual calculation or label formatting
required.

### Different dimension pairs

Explore PC2 vs PC3:

```{r}
#| label: pca-dims
#| fig.width: 7
#| fig.height: 5
do_PcaPlot(pbmc3k, dims = c(2, 3))
```

## The unified entry point

`do_DimPlot()` routes to the appropriate handler based on the `reduction`
argument:

| `reduction` | Routes to | Axis labels |
|-------------|-----------|-------------|
| `"umap"` | `do_UmapPlot()` | UMAP1, UMAP2 |
| `"pca"` | `do_PcaPlot()` | PC1 (X.X%), PC2 (X.X%) |
| other | `.do_DimPlot_internal()` | REDUCTION1, REDUCTION2 |

```{r}
#| label: dimplot-routing
#| fig.width: 14
#| fig.height: 5
p1 <- do_DimPlot(pbmc3k, reduction = "umap") + ggtitle("reduction = 'umap'")
p2 <- do_DimPlot(pbmc3k, reduction = "pca") + ggtitle("reduction = 'pca'")

p1 + p2
```

Use `do_DimPlot()` when writing general-purpose code that should work with any
reduction. Use the specific functions (`do_UmapPlot()`, `do_PcaPlot()`) when
you know the reduction type.

## Feature expression plots

`do_FeaturePlot()` overlays gene expression on the embedding using a viridis
colour scale.

```{r}
#| label: featureplot
#| fig.width: 14
#| fig.height: 5
p1 <- do_FeaturePlot(pbmc3k, features = "CD3D")
p2 <- do_FeaturePlot(pbmc3k, features = "CD14")
p3 <- do_FeaturePlot(pbmc3k, features = "MS4A1")

p1 + p2 + p3
```

Cells are ordered by expression by default (`order = TRUE`), so high-expressing
cells are plotted on top — avoiding the common problem of rare populations being
hidden beneath low-expressing cells.

## Split-panel silhouettes

This is BadranSeq's signature visualisation feature. When you pass `split.by`,
each panel shows:

1. A **grey silhouette** of all cells (spatial context)
2. **Black borders** on all cells
3. **Coloured cells** only for the current split category

This preserves the spatial layout across panels, making it easy to see where a
subpopulation sits relative to the full dataset.

First, add a synthetic condition to demonstrate:

```{r}
#| label: split-setup
set.seed(42)
pbmc3k$condition <- sample(
  c("Control", "Treatment"),
  ncol(pbmc3k),
  replace = TRUE
)
```

Now compare Seurat's default split with BadranSeq's silhouette approach:

```{r}
#| label: split-comparison
#| fig.width: 14
#| fig.height: 5
p1 <- Seurat::DimPlot(pbmc3k, reduction = "umap", split.by = "condition") +
  ggtitle("Seurat::DimPlot(split.by)")

p1
```

```{r}
#| label: split-silhouette
#| fig.width: 14
#| fig.height: 7
do_UmapPlot(pbmc3k, split.by = "condition")
```

In the Seurat version, each panel only shows its own cells — you lose the sense
of where those cells sit in the overall structure. In BadranSeq's version, the
grey background preserves this context: you can immediately see which clusters
are enriched in each condition.

The silhouette approach works with any grouping:

```{r}
#| label: split-annotations
#| fig.width: 14
#| fig.height: 7
do_UmapPlot(pbmc3k, group.by = "seurat_annotations", split.by = "condition")
```
